# An Ecological Modeler’s Primer on R

```{r, include=FALSE, warning=F, message=F}
# data mgmt
library(tidyverse)
library(lubridate)
# visualization
library(scales)
library(viridis)
```

## Part 1

Selected commands from Part I of the Primer.

### 3.3 Writing scripts

```{r, warning=F, message=F}
r = .2
N0 = 100
N1 = N0*exp(r)
N1
N0
# no value assigned to N2
# N2
```

### 3.5 Accessing libraries

```{r, warning=F, message=F}
# install package `popbio`
package_list <- c("popbio")
# for loop to download the packages defined in "package_list"
for(package in package_list){
  if(!require(package, character.only = TRUE)){
    # install the package
    install.packages(package, repos='http://cran.us.r-project.org'); 
    # load the library
    library(package, character.only = TRUE)
  }
}

# list functions available in popbio
ls(package:popbio)

```

### 3.6 Commenting your code

```{r, warning=F, message=F}
# intrinsic rate of increase
r = .2
# initial population size
N0 = 100
# Calculate new population size after one unit of time
N1 = N0*exp(r)
# Output new population size to console
N1
```

### 3.9 Getting help

```{r, warning=F, message=F}
# help
help(eigen.analysis)

# example
# example(plot)

# brings up a list of all of the functions that are related to some general topic.
apropos("plot")

```

## Part II

### 4.3 Numbers

```{r, warning=F, message=F}
# utilize variables defined above
N1 = 100*exp(r)
# set N0 equal to 100
N0 = 100
```

### 4.4 Order of operations

```{r, warning=F, message=F}
# parentheses matter 
1000*1.06^20/12
# parentheses matter 
100*1.06*(20/12)

# see pi
pi
```

## Exercise: The diameter of the earth

```{r, warning=F, message=F}
# enter an equation to calculate the diameter of the earth
circum_mi <- 24901.55
circum_km <- circum_mi / 0.621
earth.diameter = circum_km / pi
earth.diameter
```

### 5.3 Vectors

```{r, warning=F, message=F}
# create a vector
v1 = c(234, 17, 42.5, 64)
v1
# access third element of vector
v1[3]
# vector with strings
study.areas = c("Maybell", "Poudre", "Gunnison")
```

## Exercise: Vectors 1

```{r, warning=F, message=F}
# What is the value of v1[2]?
v1[2]
# vector of ages
ages = c(22, 25, 26, 36)
# vector of names
names = c("Brynn", "Caroline", "Carolyn", "George")
# Output the value of the fourth element of v1
v1[4]
# Multiply the second element of v1 by 2. 
v1[2] * 2
# Enter v1[1:3] in the console.
v1[1:3]
# What does the : operator do?
print("the `:` operator creates a list from the value before the `:` to the value after the `:`")
```
  
## Exercise: logical operations on vectors

```{r, warning=F, message=F}
# filter v1 in various ways
v1; v1 < 200; v1[v1 < 200] 
# form a new vector from the from elements of v1 containing elements that are 
# a) greater than twenty 
a = v1[v1>20]
# b) less than 200 and greater than 20
b = v1[v1>20 & v1<200]
# c) not equal to 17
c = v1[v1!=17]
# d) equal to 17 or equal to 42.5
c = v1[v1==17 | v1==42.5]
xx = c(17, 42.5)
c = v1[v1 %in% xx]
```

## Exercise: vector functions

```{r, warning=F, message=F}
# Find the mean, max, minimum, range, and variance of v1.
print(paste("The mean of v1 is", mean(v1)))
print(paste("The max of v1 is", max(v1)))
print(paste("The min of v1 is", min(v1)))
range(v1)
print(paste("The var of v1 is", var(v1)))

# Compose another vector v2 containing 4 numeric values of your choice. 
v2 = c(20:24)

# Now, execute the commands min(v1,v2)
min(v1,v2)
pmin(v1,v2)

print("pmin returns the minimum of all the values present in their argument")

```

## Exercise: more vector functions

```{r, warning=F, message=F}
d = numeric(5)
d
(d + 1) /3
length(d)
print("d is a numeric vector of length 5 with all values set to 0; math can be performed for each value in the vector")
# declare undefined vector
v = NULL
v[8] = 254
v[2] = 10
v
typeof(v)
v[12] = 13.4
v
typeof(v)
print("initially v was a vector that had double numeric values, when the value with the decimal was included R printed the double value with the decimal place")
```

### 5.3.6 Creating special vectors

```{r, warning=F, message=F}
v = seq(1,10)
rep(c("apple", "bookshelf"),5)
```

## Exercise: Creating a matrix

```{r, warning=F, message=F}
# Create and output a matrix named B that contains seven
#  columns and 5 rows with all elements containing the value 7
mtrx_dta = c(7)
B = matrix(mtrx_dta, nrow=5, ncol=7, byrow=TRUE)
B
# specifies all of the elements in the second column of the matrix
B[,2]
# specifies all of the elements of the fourth row of the matrix A
B[4,]
```

## Exercise: matrices

```{r, warning=F, message=F}
# Enter the data in the Excel example above into an matrix called M.
# 1. Sets up the matrix M and fills it with zeros using the matrix() statement.
M = matrix(data = NA, nrow = 5, ncol = 2)
# 2. Enters the appropriate values in the matrix row by row Hint: to enter the first row
M[1,] = c(1995, 10.0)
M[2,] = c(1996, 12.5)
M[3,] = c(1997, 15.6)
M[4,] = c(1998, 19.5)
M[5,] = c(1999, 24.4)
# 3. Outputs the entire matrix
M
# 4. Outputs each row
M[1,]
M[2,]
M[3,]
M[4,]
M[5,]
# 5. Outputs each column.
M[,1]
M[,2]
# 6. Finds the largest value in the matrix
max(M)
# 7. Finds the average of the second column
mean(M[,2])
```

## Exercise: matrices, continued.

```{r, warning=F, message=F}
# 1. Find the all of the values of column 2 of M that are less than 19.
M[ M[,2] < 19 ,]

# 2. Find the values of the indices of column 2 of M that are less than 19.
which(M[,2] < 19)

# 3. Find and output the row that contains the minimum value contained in column 2.
min_val <- min(M[,2])
M[ M[,2] == min_val, ]

# 4. Now, to illustrate a particularly devious error, enter the following at the console: M,
# M[1], M[10], mean(M[1]), mean(M[,1]). Discuss what is going on here with your
# lab mates. What problems that can arise if you leave out the comma when you intend
# to index a row or column? When you enter a single subscript (with no comma) for
# a two dimensional matrix, how is R treating the matrix? Hint: enter the commands
# length(M), length(M[1,]), length(M[,1]at the console. This is particularly important to understand because it is a source of potential errors that are hard to detect,
# illustrating that flexible languages like R give you tremendous power—but with that
# power comes the responsibility to understand what you are doing. Don’t go forward
# until you really understand this part of the exercise.
```

## Exercise: matrices

```{r, warning=F, message=F}
d = numeric(5)
d
(d + 1) /3
length(d)
print("d is a numeric vector of length 5 with all values set to 0; math can be performed for each value in the vector")
```